\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage[hidelinks]{hyperref}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{listings}
\usepackage{xcolor}

\lstset{
  basicstyle=\ttfamily\small,
  breaklines=true,
  frame=single,
  backgroundcolor=\color{gray!8},
}

\title{\textbf{FluctlightDB: A Memory Model of Data for AI Agents}\\[0.4em]
\large A brain-native database engine for long-term agent memory}
\author{Ganesh S\\ \small Independent Researcher \quad \texttt{voxmastery@gmail.com}\\[0.2em]
\small ORCID: \href{https://orcid.org/0009-0006-7758-4114}{0009-0006-7758-4114}}
\date{June 2025}

\begin{document}
\maketitle

\begin{abstract}
For fifty years, data systems have answered two questions. The relational model
asked \emph{which records match a predicate}; the vector model asked \emph{which
vectors lie nearest a query}. Neither was built to answer the question an
autonomous agent asks every time it wakes up: \emph{what have I learned, and what
of it can I trust?} We argue that long-term agent memory is not an application
built on top of a database---it is a \textbf{third data model}, with its own write
semantics (encoding, separation, consolidation, provenance) and its own read
semantics (cue-driven activation across a linked memory graph). We present
\textbf{FluctlightDB}, an embedded, brain-native database engine that implements
this model behind two primitives, \texttt{experience()} and \texttt{activate()}.
On the official LoCoMo long-conversation benchmark (10 conversations,
1{,}982 gold evidence spans), FluctlightDB's hybrid index recalls \textbf{98.1\%}
of gold evidence---warm and cold-start identical. On BEIR SciFact it matches a
tuned Chroma + MiniLM baseline on nDCG@10 at equal latency, while agent mode
improves Recall@100. On an agent-specific suite (FAMB) it scores 97--98\% macro
across paraphrase recall, provenance ranking, persistence, and write-time
separation. The engine, harnesses, and frozen metrics are released under the MIT
license. We claim no new neuroscience and no new transformer; we claim a missing
layer of the data stack, and an engine that fills it.
\end{abstract}

\section{Introduction}
Every generation of software gets the database it deserves. Business records gave
us the relational model~\cite{codd1970relational} and SQL. Embeddings gave us
vector databases and approximate nearest-neighbor search. Agents---programs that
act, observe, and persist across sessions---have so far been handed \emph{neither}.
They are stateless between runs unless a developer hand-assembles a session store,
a vector index, a deduplicator, a trust policy, and glue code to keep them
consistent. The result is that ``memory'' for agents lives in application code,
re-implemented badly in every project, with no shared abstraction and no
benchmark to hold it accountable.

This paper makes a deliberately large claim and then defends it with measurements:
\textbf{agent memory deserves its own database engine, not a wrapper around someone
else's.} A relational engine is the wrong abstraction because memory is not a set
of typed rows joined by keys. A vector engine is the wrong abstraction because
recall is not cosine similarity alone---a human, and an agent, retrieve a fact
because it was \emph{learned}, \emph{linked} to a context, and \emph{trusted},
not merely because its embedding is close. What is missing is a model whose unit
is the \emph{episode}---an experience with content, context, salience, and
provenance---and whose query is a \emph{cue} that spreads through associations.

\textbf{Contributions.}
\begin{itemize}
  \item \textbf{A data model.} We define agent memory as a first-class model of
  data with explicit write semantics (separation, encoding, consolidation,
  provenance) and read semantics (cue activation over a memory graph), distinct
  from the relational and vector models.
  \item \textbf{An engine.} FluctlightDB, an embedded Rust engine exposing
  \texttt{experience()} / \texttt{activate()} / \texttt{checkpoint()}---one
  durable store per agent, with no external services.
  \item \textbf{Evidence.} \textbf{98.1\%} evidence recall on full LoCoMo,
  parity with a tuned vector baseline on BEIR SciFact, and 97--98\% macro on an
  agent-specific benchmark.
  \item \textbf{Reproducibility.} Open harnesses and frozen result JSON so every
  number here can be re-run with one command.
\end{itemize}

\section{The Third Data Model}
\subsection{Why rows are not memory}
The relational model stores facts whose schema is known in advance and whose
truth is uniform. Agent memory is the opposite: heterogeneous, arriving out of
order, often contradictory (a chat claim vs.\ a ledger entry), and valuable
precisely \emph{because} of where it came from. SQL has no native notion of
provenance-weighted recall or of an experience that should be down-weighted
because a more trusted source disagrees.

\subsection{Why nearest-neighbor is not recall}
Vector search answers ``what is similar.'' Memory answers ``what is relevant given
who I am and what I was doing.'' Two memories with distant embeddings can be the
right answer because they were co-activated during the same episode; a near
embedding can be the wrong answer because it is an unverified rumor. Pure ANN has
no place to encode association strength, salience decay, or trust.

\subsection{The model}
We define a memory store as a set of \emph{engrams}. Each engram carries content,
an encoding context, a salience weight, optional provenance, and edges to
co-activated engrams. The write operation \texttt{experience} performs pattern
\emph{separation} (near-duplicates are gated, not blindly appended), encodes the
engram, registers its semantic vector, and wires graph edges. The read operation
\texttt{activate} takes a cue, seeds both lexical and semantic indexes, spreads
activation through the graph, fuses the scores, and applies provenance boosts so
verified sources outrank chat. Consolidation replays and compacts the store
offline, the way sleep consolidates biological memory. This is the entire
contract; the neuroscience vocabulary is explanatory, not required to use it.

\begin{table}[h]
\centering
\caption{Brain-native primitives vs.\ relational and vector abstractions.}
\begin{tabular}{p{2.2cm}p{4.2cm}p{4.8cm}}
\toprule
\textbf{Primitive} & \textbf{Role in FluctlightDB} & \textbf{Relational / vector analogue} \\
\midrule
Engram & Unit of memory: content, context, salience, provenance, edges & Row; or embedded chunk \\
\texttt{experience()} & Write: separation, encode, index, wire co-activation & \texttt{INSERT} / upsert \\
\texttt{activate(cue)} & Read: lexical + semantic seed, graph spread, fusion, trust boost & \texttt{SELECT} / ANN top-$k$ \\
Consolidation & Offline replay and compaction (\texttt{checkpoint()}) & Vacuum / reindex only \\
Provenance & Verified sources outrank unverified chat at recall & No native type \\
\bottomrule
\end{tabular}
\end{table}

\section{System Design}
\subsection{Embedded, one store per agent}
Like SQLite, FluctlightDB is a library, not a server: each agent owns a directory
on disk and \texttt{checkpoint()} commits state. There is nothing to provision and
nothing to keep in sync. A thin HTTP server is provided for multi-agent
deployments but is not required.

\subsection{Write path}
Incoming \texttt{experience} episodes pass through optional separation gating,
dentate-style encoding into engrams, semantic vector registration, and graph
wiring. A fast-ingest mode skips heavy graph work for bulk indexing.

\subsection{Read path}
\texttt{activate(cue)} seeds lexical (BM25) and semantic (vector) indexes, spreads
activation through the memory graph, fuses scores, and applies provenance boosts.
For conversational RAG we merge vector top-$k$ with lexical seeds---\emph{hybrid
retrieval}---before ranking. A vector-fast mode provides Chroma-class cosine
top-$k$ for IR comparisons.

\subsection{Two explicit modes}
\texttt{connect()} is the full episodic engine (separation, graph, provenance) for
live agents. \texttt{connect\_index()} is the bulk semantic path for RAG backfills
and IR benchmarks. The modes share one engine and one file format.

\section{Evaluation}
All experiments use \texttt{all-MiniLM-L6-v2} (ONNX, CPU) unless noted. Every
number below is reproduced by a script in \texttt{benchmarks/} and frozen in
\texttt{benchmarks/results/2025-06-22.json}.

\subsection{BEIR SciFact: parity with a tuned vector baseline}
We score standard IR metrics via \texttt{pytrec\_eval} against official qrels,
using identical embeddings for FluctlightDB and Chroma. Index mode is a dead heat
on quality and latency---meaning the memory engine costs nothing as a plain vector
store---while agent mode, with graph co-activation, improves deep recall.

\begin{table}[h]
\centering
\caption{BEIR SciFact (shared MiniLM embeddings).}
\begin{tabular}{lrrrr}
\toprule
System & nDCG@10 & R@10 & R@100 & Query (ms) \\
\midrule
Chroma & 0.645 & 0.783 & 0.925 & 5--7 \\
FluctlightDB (index) & 0.645 & 0.783 & 0.925 & 5--7 \\
FluctlightDB (agent) & \textbf{0.651} & \textbf{0.790} & \textbf{0.941} & 15 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{LoCoMo: 98.1\% evidence recall on the full set}
LoCoMo~\cite{maharana2024locomo} stresses very long, multi-session dialogue. We
report the official evidence-recall metric: the fraction of gold \texttt{dia\_id}
evidence spans present in retrieved context. Configuration: \texttt{connect\_index()},
ingest of dialog \emph{and} observations, $k{=}150$, hybrid vector+BM25, two CPU
threads. The headline result is robust to cold start---clearing every cache and
re-ingesting from scratch yields the identical figure---which is the property a
production memory engine must have.

\begin{table}[h]
\centering
\caption{LoCoMo full set: 10 conversations, 1{,}982 gold evidence spans.}
\begin{tabular}{lrr}
\toprule
Metric & Warm & Cold (caches cleared) \\
\midrule
Mean evidence recall & \textbf{98.1\%} & \textbf{98.1\%} \\
All evidence in context & 97.1\% & 97.1\% \\
Evidence hits & 1925/1982 & 1925/1982 \\
Wall time (s) & 271 & 335 \\
\bottomrule
\end{tabular}
\end{table}

We deliberately separate \emph{retrieval} from \emph{generation}. A verbatim
answer-in-context proxy sits near 38\%, but this measures string overlap, not
correctness---LoCoMo gold answers are usually inferred facts (dates, counts), not
quoted spans. A 50-question end-to-end pilot with a hosted reader LLM produced
23.5\% category F1 at 99.5\% retrieval on the same slice; full end-to-end QA awaits
sufficient inference quota and is reported as future work. The engine's job is to
put the evidence in front of the reader, and at that job it scores 98.1\%.

\subsection{FAMB: the agent-specific suite}
Generic IR does not test what agents need. FAMB measures paraphrase recall@1
(real cues never match stored wording), provenance top-1 (verified ledger beats
chat claim), persistence (recall survives checkpoint + reopen), confusion ingest
(near-duplicates do not block new facts), and determinism. Index-mode macro is
\textbf{98\%}; agent-mode macro is \textbf{97\%}.

\subsection{LongMemEval-S}
LongMemEval~\cite{wu2024longmemeval} evaluates six multi-session abilities with
end-to-end LLM QA. Our retrieval-only harness measures answer-in-recall@8. A
20-question pilot reached 70\%; the full 500-question run is CPU-bound
($\sim$30\,s/question for per-question ingest with CPU embeddings) and is deferred
to a throttled batch rather than reported prematurely.

\section{Discussion}
\textbf{The engine is the contribution, not the reader.} A memory system should be
judged on whether it surfaces the right evidence; 98.1\% on full LoCoMo says it
does. End-to-end QA additionally depends on the language model and the prompt
budget, and we are careful not to launder a retrieval win into a generation claim.

\textbf{In production.} Beyond benchmarks, FluctlightDB backs a continuously
running operational agent in a deployed production system, persisting
cross-session state for an autonomous service rather than a research fixture---the
durability and cold-start guarantees above are properties we rely on, not only
ones we measured.

\textbf{Hybrid retrieval matters.} Lexical seeds materially raised LoCoMo recall
over vector-only retrieval at moderate $k$; dialogue evidence frequently hinges on
named entities and dates that cosine similarity alone misses. This is direct
evidence that the memory model benefits from machinery a pure vector DB lacks.

\textbf{Limitations.} Per-question ingest on LongMemEval is CPU-heavy under a CPU
embedder. We have not yet evaluated multi-tenant isolation at scale, ANN beyond
$10^5$ memories, or full LLM QA against published Mem0/Zep baselines. These are
engineering and quota problems, not model problems, and are the natural next steps.

\section{Related Work}
\subsection{Vector databases}
Systems such as Chroma, Qdrant, and FAISS~\cite{thakur2021beir} optimize approximate
nearest-neighbor search. They answer \emph{similarity}, not \emph{memory}: there is
no native episode, provenance-weighted recall, write-time separation, or
cue-driven spreading activation. FluctlightDB's \texttt{connect\_index()} mode
deliberately exposes a vector-fast path so we can measure parity on BEIR; agent
mode adds graph co-activation on top of the same store.

\subsection{Agent memory SDKs and layers}
\textbf{Mem0}~\cite{chhikara2025mem0} is the closest contemporary system: it
extracts, consolidates, and retrieves salient facts from dialogue, with an optional
graph variant (Mem0$^{\tiny g}$) for relational structure among entities. It
reports strong LoCoMo numbers using \emph{LLM-as-a-judge end-to-end QA}---a
different metric from the official evidence-recall protocol we adopt for retrieval.
\textbf{Zep}~\cite{zep2024} provides temporal knowledge-graph memory for
assistants, also evaluated with reader-LLM QA on LoCoMo. \textbf{Cognee}~\cite{cognee2024}
pipelines vector and graph storage for agent RAG. \textbf{MemGPT}~\cite{packer2023memgpt}
(Letta) treats context as an OS resource with explicit memory tiers but does not
define a standalone embedded memory \emph{engine contract}. \textbf{HippoRAG}~\cite{gutierrez2024hipporag}
draws on associative-retrieval neuroscience for multi-hop QA over a graph index.
These systems are valuable \emph{memory layers}: they orchestrate extraction,
summarization, and retrieval over general-purpose backends. We position
FluctlightDB as the missing \emph{engine} beneath such layers---where the native
operations are \texttt{experience} and \texttt{activate}, not SQL or
\texttt{vector\_search}.

\subsection{Positioning}
To our knowledge, no prior work argues that long-term agent memory is a third data
model---peer to the relational~\cite{codd1970relational} and vector models---with
engine-level write and read semantics implemented in an embedded store. Enterprise
databases adding agent features (e.g.\ Oracle AI Database unified memory) retrofit
memory onto existing engines rather than defining episodes and cues as first-class
types. Table~\ref{tab:landscape} summarizes the landscape; direct head-to-head
retrieval on LoCoMo evidence recall against Mem0/Zep is future work (different
published metrics today).

\begin{table}[h]
\centering
\caption{Landscape: memory \emph{layers} vs.\ FluctlightDB as an embedded \emph{engine}. LoCoMo figures are \emph{not} directly comparable across rows unless the metric column matches.}
\label{tab:landscape}
\small
\begin{tabular}{p{2.0cm}p{1.6cm}p{2.4cm}p{2.6cm}p{2.2cm}}
\toprule
\textbf{System} & \textbf{Kind} & \textbf{Native contract} & \textbf{LoCoMo (cited)} & \textbf{Metric type} \\
\midrule
Mem0 / Mem0$^{\tiny g}$ & SDK + backend & Extract / consolidate / retrieve facts & $\sim$92\%+ LLM-J & End-to-end QA \\
Zep & Managed layer & Temporal KG + summaries & $\sim$75\% LLM-J & End-to-end QA \\
Cognee & Pipeline & Graph + vector ETL & --- & Task-specific \\
MemGPT / Letta & Agent OS & Context tiers / blocks & --- & Session QA \\
HippoRAG & Graph RAG & Associative graph retrieval & --- & Multi-hop QA \\
Chroma / Qdrant & Vector DB & ANN top-$k$ & --- & IR / similarity \\
FluctlightDB & Engine & \texttt{experience} / \texttt{activate} & \textbf{98.1\%} & Evidence recall \\
\bottomrule
\end{tabular}
\end{table}

Benchmarks LoCoMo~\cite{maharana2024locomo} and LongMemEval~\cite{wu2024longmemeval}
measure long-horizon memory; BEIR~\cite{thakur2021beir} anchors IR credibility.
Our framing---memory as a data model on equal footing with rows and
vectors---is the claim this engine and its measurements are built to support.

\section{Conclusion}
The relational model gave applications a database for \emph{facts}; the vector
model gave search a database for \emph{similarity}. Autonomous agents need a
database for \emph{memory}, and it should be as boring to adopt and as rigorous to
trust as SQLite. FluctlightDB is our argument that this engine can exist today: it
matches vector baselines where they are strong, wins where memory semantics
matter, and recalls 98.1\% of gold evidence on the hardest public long-conversation
benchmark. We release everything needed to reproduce and contest these claims.

\section*{Artifacts}
Repository: FluctlightDB (MIT). Harnesses: \texttt{benchmarks/locomo\_eval.py},
\texttt{benchmarks/longmemeval\_bench.py}, \texttt{benchmarks/beir\_bench.py},
\texttt{benchmarks/agent\_memory\_bench.py}. Frozen metrics:
\texttt{benchmarks/results/2025-06-22.json}. Live draft viewer:
\url{https://github.com/voxmastery/FluctlightDB/tree/main/papers/arxiv-v1}.

\bibliographystyle{plain}
\bibliography{references}
\end{document}
